Shapes analysis for time series.

### 2024-05-15 NEURIPS24/neurips_2024_to_submit.tex
polish intro

\begin{abstract}
    %TO CHANGE]
    Analyzing inter-individual variability of physiological functions is particularly appealing in medical and biological contexts to describe or quantify health conditions. Such analysis can be done by comparing individuals to a reference one with time series as biomedical data. 
This paper introduces an unsupervised representation learning (URL) algorithm for time series tailored to inter-individual studies. The idea is to represent time series as deformations of a reference time series. The deformations are diffeomorphisms parameterized and learned by our method called TS-LDDMM. Once the deformations and the reference time series are learned, the vector representations of individual time series are given by the parametrization of their corresponding deformation. At the crossroads between URL for time series and shape analysis, the proposed algorithm handles irregularly sampled multivariate time series of variable lengths and provides shape-based representations of temporal data.
 In this work, we establish a representation theorem for the graph of a time series and derive its consequences on the LDDMM framework. We showcase the advantages of our representation compared to existing methods using synthetic data and real-world examples motivated by biomedical applications.
\end{abstract}

>>>>

\begin{abstract}
Analyzing inter-individual variability of physiological functions is particularly appealing in medical and biological contexts to describe or quantify health conditions. Such analysis can be done by comparing individuals to a reference one with time series as biomedical data.
This paper introduces an unsupervised representation learning (URL) algorithm for time series tailored to inter-individual studies. The idea is to represent time series as deformations of a reference time series. The deformations are diffeomorphisms parameterized and learned by our method called TS-LDDMM. Once the deformations and the reference time series are learned, the vector representations of individual time series are given by the parametrization of their corresponding deformation. At the crossroads between URL for time series and shape analysis, the proposed algorithm handles irregularly sampled multivariate time series of variable lengths and provides shape-based representations of temporal data.
In this work, we establish a representation theorem for the graph of a time series and derive its consequences on the LDDMM framework. We showcase the advantages of our representation compared to existing methods using synthetic data and real-world examples motivated by biomedical applications.
\end{abstract}

### 2024-05-15 NEURIPS24/paper.tex
polish intro

>>>>

\thi{
  Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}. More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
}

\thi{
  Studying shape differences between time series related to biological mechanisms is a common practice in physiology to characterize healthy and pathological functioning CITE. For instance, the shapes of heartbeats in electrocardiograms are discriminant for some cardiovascular pathologies CITE. Several approaches have been proposed for such comparison. Some employ shape-based similarity measures between time series \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k,ye2009time},  others embed time series as vectors of predefined features CITE, and, with the rise of deep neural networks, unsupervised learning representation of time series~\cite{trirat2024universal,meng2023unsupervised} has shown to be a valuable approach CITE notably with contrastive learning~\cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised}. However, shape-based representation of time series within cohorts~\cite{niennattrakul2007inaccuracies,shirato2023identifying} remains an open problem in URL.
}

\thi{
  Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$. Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$ (ie $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ ). The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS). The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
  }

\thi{
  However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}. Our contributions arise from this observation: we specify the class of diffeomorphisms to consider and show how to learn them. This change is fruitful in representing transformations of time series graphs as illustrated in \Cref{fig:transport}.
}

\thi{
  Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
 \end{itemize}
}
